\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization \thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department of Computer Science, Brigham Young University, 3361 TMCB, Provo, UT 84602 (phone: 801-422-8717; email: \{mjg82,a,k\}@cs.byu.edu).}

### 2010-02-25 honors_thesis/thesis/paper.tex
Final draft of thesis submitted to be bound

\Author{Matthew Gardner}
\Advisor{Kevin Seppi}
\HonorsRep{}
\Abstract{
Particle swarm optimization (PSO) has previously been parallelized only by 
adding more particles to the swarm or by parallelizing the evaluation of the
objective function.  However, some functions are better optimized with more
iterations and fewer particles.  Accordingly, we take inspiration from 
speculative execution commonly performed in modern processors and propose what
we call speculative evaluation in PSO.  Future positions of the particles are
speculated and evaluated in parallel with current positions, performing two
iterations of PSO at once.

@@

There are many ways of defining the neighbors of a particle.  Different
neighborhood topologies have varying performance on different kinds of
functions, as would be expected by the No Free Lunch
theorem~\cite{wolpert-tec97}.  An in-depth discussion of topologies in PSO is
beyond the scope of this paper.  The two most common topologies used in the
literature are the Ring topology and the Complete
topology~\cite{bratton-sis07}.  In this paper we use those topologies, as well
as a parallel adaptation of the Complete topology called Random that has been
shown to simulate the behavior of Complete with far less
communication~\cite{mcnabb-cec09}.  Both the Ring and the Random topologies can
be defined as having $k$ neighbors; we use 2 neighbors for all of our
experiments.

@@

To enumerate all of the possible next positions of a particle, one must simply
evaluate the motion equations for all possible combinations of personal best
and neighborhood best updates that a particle could see.  This produces
$2(n+1)$ speculative positions to be evaluated, where $n$ is the number of
neighbors the particle has.  If a particle is also its own neighbor, as is most
often the case, one of the $2(n+1)$ speculative positions can be eliminated, as
a particle cannot be the source of its $\nbest$ update while not updating its
$\pbest$.  Thus we have $2(n+1)-1$ evaluations.  Table~\ref{tab:evals} lists
all possible combinations of updates that can be seen by a particle with two
neighbors other than itself, which is the case in all of the topologies we use.
If all of these speculative positions are evaluated in parallel along with the
original, non-speculative position of the particle at time $t-1$, then one of
the speculative positions has to be the correct one, and two iterations have
been evaluated at once.

@@

Because $2(n+1)-1$ speculative evaluations must be performed for each particle,
the choice of topology is an important one.  The use of the Complete topology,
where every particle is a neighbor to every other particle, would require
$O(p^2)$ speculative evaluations per iteration, where $p$ is the swarm size.
Clearly it is much more desirable to have a sparse topology, where $O(np)$ is
much smaller than $O(p^2)$.  However, some functions are better optimized with
the Complete topology and the quick spread of information it entails than with
sparse topologies.  Accordingly, we use the Random topology described
in~\cite{mcnabb-cec09}, which has been shown to effectively simulate the
Complete topology.  In that topology, 2 neighbors are chosen at random from the
swarm for each particle at each iteration.  Because a particle is also always
its own neighbor, $n$ is $3$ in that case.  In~\sec{results} we report the
results for performing speculative evaluation using both the Ring topology
(where $n$ again is $3$) and the Random topology on a number of common
benchmark functions.

It is not trivial in some parallel architectures to determine which speculative
position was the correct next position of each particle.  First we discuss the
relatively easy case of a centralized parallel PSO algorithm.  In such an
architecture, the master keeps track of all necessary information with only
trivial message passing needed.  Then we discuss the more complicated case of a
distributed algorithm, where each particle is on its own and needs to send and
receive messages to and from other particles.  Finally we discuss the further
complications of a dynamic topology, where a particle's neighbors change from
one iteration to another.

@@

Upon picking the correct branch for each particle and updating the children's
personal best and neighborhood best value (from iteration $t$), the result is
the set $\nonbest{\pset}_{t+1}$, as the particles are now no longer
speculative.  What remains is to update the neighborhood best of those
particles from their neighbors (from iteration $t+1$), as above, to obtain
$\pset_{t+1}$.  That set of particles can subsequently be used to produce the
sets $\noeval{\pset}_{t+2}$ and $\noeval{\sset}_{t+3}$ (steps 1 and 2 in
\alg{centralized}), and the process repeats itself.

@@

We have already described how to use a set of messages to obtain $\p_{t+1}$.
The process is exactly the same to produce each $\n_{t+1}$, requiring the same
messages, except for the neighbor particles instead of the particle itself.  
\fig{messages5} shows graphically how $\noeval{\sset}_{t+3}$ is produced.

@@

Performing speculative evaluation in PSO with a dynamic topology raises a few
issues of its own.  First, with neighbors changing each iteration, messages
that processors pass to their neighbors need to be sent to the correct
neighbors for each iteration.  In centralized algorithms and in distributed
algorithms with several rounds of message passing (the first method described
above), nothing needs to be changed.  However, in the case where processors
reconstruct information about neighboring particles from messages, those
messages need to be sent to the correct neighbors.  A particle cannot simply
send messages to its neighbors' neighbors' neighbors---it needs to send
messages to its iteration $t$ neighbors' iteration $t+1$ neighbors, and so on.
For every neighbor outward information is sent, the iteration also needs to be
incremented, as information about neighbors' neighbors is used during iteration
$t+1$, and information about neighbors' neighbors' neighbors is used to
reconstruct information about iteration $t+2$.  Also, this method of message
passing again requires the use of random seeds if the topology is random, so
that each processor computes the same neighbors for a particle as all other
processors.

Also, an issue common to all methods of performing speculative evaluation is
that of getting available information from neighbors before generating
speculative children.  In a static topology, at iteration $t$ a particle
already has all of the information about the positions of its neighbors during
iterations $1$ through $t-1$.  If the neighbor finds a better position at 
iteration $t$, the particle updates its neighborhood best, but if it does not,
it still has its old neighborhood best from its neighbors for all previous 
iterations.

In a dynamic topology, a particle might not have information about the previous
positions of its neighbors at iteration $t$.  That means that its new
neighborhood best could come not only from its neighbors' positions at
iteration $t$, but also from their personal best from iteration $t-1$, as
neighbors' personal bests are what are used to update a particle's neighborhood
best.  That creates a problem for speculative evaluation---there are
potentially more than $2(n+1)-1$ possible next positions, increasing the amount
of work that must be done to perform the second iteration at the same time as
the first.

This is easily fixed by updating $\noeval{\p}_{t+2}$ with the currently
available information about $\noeval{\nset}_{t+2}$ before producing
$\noeval{\sset}_{t+3}$.  If a particle $\noeval{\p}_{t+2}$ updates its 
neighborhood best with the personal bests of $\noeval{\nset}_{t+2}$ before
calculating the next possible positions for $\noeval{\sset}_{t+3}$, there are
still only $2(n+1)-1$ possible next positions, and the problem is avoided.

@@

In performing speculative evaluation as we have described it, $2(n+1)-1$
speculative evaluations are done per particle, while all but one of them are
completely ignored.  It seems reasonable to try to make use of those
evaluations instead of throwing them away.  Making use of them changes the
behavior of PSO, instead of reproducing it exactly as the above method
explains, but the change could be an improvement.

@@

In our experiments we compared our speculative PSO algorithm to the standard
PSO algorithm.  At each iteration of the algorithms, we use one processor to
perform one function evaluation for one particle, be it speculative or
non-speculative.  The speculative algorithm actually performs two iterations of
PSO at each ``iteration,'' so we instead call each ``iteration'' a round of
evaluations.  For benchmark functions with very fast evaluations this may not
be the best use of parallelism in PSO.  But for many real-world applications,
the objective function takes on the order of minutes or more to evaluate; in
such cases our assumption is perfectly reasonable.

@@

Next we look at the function Griewank, defined by the equation $f(\Vec{x}) =
\frac{1}{4000} \sum_{i=1}^D x_i^2 - \Pi_{i=1}^D \cos\left(\frac{x_i}{\sqrt{i}}
\right) + 1$.  It is generally recommended to use the Ring topology when
optimizing the Griewank function, as Complete is prone to premature convergence
on a local optima.  We show results in \fig{griewank1} for swarms of size 30
and 240 using the Ring topology. One can see in the figure that both SEPSO
algorithms initially outperformed PSO, but got stuck in a local optima and were
eventually outperformed themselves.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank1.eps}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations.  Speculative algorithms have a swarm of 30 particles and do two
  iterations per round, while standard PSO algorithms use 240 processors and do
  one iteration per round.}
  \label{fig:griewank1}
\end{figure}

\fig{griewank1} is slightly deceptive, however, as with Griewank some of the
runs found the global optimum, 0, while some did not.  A more accurate plot for
this function is one showing what percentage of the runs have found the global
optimum at each iteration.  \fig{griewank2} shows such a plot.  A swarm of size
30 performing the PSO algorithm gets stuck a little less than half of the time,
while 240 particles provides enough exploration to find the global optimum
every time.  Surprisingly, using the SEPSO Pick Best algorithm with 30
particles greatly improves the chance of success with Griewank, while also
finding the global optimum in far fewer rounds of evaluations than standard
PSO.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank2.eps}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations.  Instead of showing average function value, we show the
  percentage of runs that have found the global optimum by each iteration.}
  \label{fig:griewank2}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{rastrigin.eps}
  \caption{Function Rastrigin with a swarm that uses 240 processors per round
  of evaluations.  Speculative algorithms have a swarm of 30 particles and do
  two iterations per round, while standard PSO algorithms use 240 processors
  and do one iteration per round.}
  \label{fig:rastrigin}
\end{figure}

@@

We have described how parallel implementations of particle swarm optimization
can be modified to allow additional processors to increase the number of
iterations of the algorithm performed, instead of merely adding more particles
to the swarm.  We have detailed several different implementations of the
algorithm, depending on the parallel PSO framework being used.  Using our
modifications, the original PSO algorithm is exactly reproduced two iterations
at a time.  This technique requires more function evaluations per iteration
than regular PSO, but for some functions still performs better when run in a
parallel environment.

We have also described a method for making use of extra speculative evaluations
that performs incredibly well on some functions.

@@

There are some functions for which very little exploration needs to be done;
Sphere is an example of such a function, as would be a function where the
general area of the optimal point is known, but the exact point needs to be
found.  For such functions the best use of processors is to have a small swarm
performing speculative evaluation with our second method, where all speculative
evaluations are used.

There are other functions for which there is a point where the algorithm has
done ``enough'' exploration and consistently finds the global optimum.  Adding
additional particles to the swarm is akin to increasing its capacity for
exploration.  We have seen that for a 20-dimensional Griewank function, a
number of particles greater than 30 but less than 240 has ``enough''
exploration and finds the optimal value every time.  For these functions, one
should use processors to increase the swarm size until enough exploration has
been reached, then use additional processors to increase the number of
iterations performed.

There are still other functions for which it seems there is never enough
exploration, such as the Rastrigin function.  It has been shown that up to 4000
particles there is no point at which ``enough'' exploration has been
done~\cite{mcnabb-cec09}.  For those functions the correct use of processors
cannot be easily prescribed and depends largely on the desired outcome.  It is
likely that the best thing to do with those functions is to use all processors
to add particles to the swarm and do no speculative evaluation.

>>>>

\Author{Matthew Gardner}
\Advisor{Kevin Seppi}
\HonorsRep{Bruce Collings}
\Abstract{
Particle swarm optimization (PSO) has previously been parallelized only by 
adding more particles to the swarm or by parallelizing the evaluation of the
objective function.  However, some functions are better optimized with more
iterations and fewer particles.  Accordingly, we take inspiration from 
speculative execution commonly performed in modern processors and propose what
we call speculative evaluation in PSO.  Future positions of the particles are
speculated and evaluated in parallel with current positions, performing two
iterations of PSO at once.

@@

There are many ways of defining the neighbors of a particle.  Different
neighborhood topologies have varying performance on different kinds of
functions, as would be expected by the No Free Lunch
theorem~\cite{wolpert-tec97}.  An in-depth discussion of topologies in PSO is
beyond the scope of this paper.  The two most common topologies used in the
literature are the Ring topology and the Complete
topology~\cite{bratton-sis07}.  In the Ring topology, each particle has one
neighbor to either side of it; in the Complete topology, every particle is a
neighbor to every other particle.  In all topologies, a particle is also a
neighbor to itself, in that its own position is considered when updating the
particle's $\nbest$.  Thus with $p$ particles, using the Ring topology each
particle $i$ has three neighbors: $i-1$, $i$, and $i+1$.  With the Complete
topology, each particle has $p$ neighbors.  In this paper we use these
topologies, as well as a parallel adaptation of the Complete topology, called
Random, that has been shown to simulate the behavior of Complete with far less
communication~\cite{mcnabb-cec09}.  In the Random topology, each particle
randomly picks two other particles to share information with at each iteration,
along with itself.  Thus in both the Ring and the Random topologies, all
particles have 3 neighbors.

@@

To enumerate all of the possible next positions of a particle, one must simply
evaluate the motion equations for all possible combinations of personal best
and neighborhood best updates that a particle could see.  This produces
$2(n+1)$ speculative positions to be evaluated, where $n$ is the number of
neighbors the particle has.  Because a particle is also its own neighbor, one
of the $2(n+1)$ speculative positions can be eliminated, as a particle cannot
be the source of its $\nbest$ update while not updating its $\pbest$.  Thus we
have $2(n+1)-1$, or $2n+1$, evaluations.  Table~\ref{tab:evals} lists all
possible combinations of updates that can be seen by a particle with two
neighbors other than itself, which is the case in all of the topologies we use.
If all of these speculative positions are evaluated in parallel along with the
original, non-speculative position of the particle at time $t-1$, then one of
the speculative positions has to be the correct one, and two iterations have
been evaluated at once.

@@

Because $2n+1$ speculative evaluations must be performed for each particle, the
choice of topology is an important one.  The use of the Complete topology,
where every particle is a neighbor to every other particle, would require
$O(p^2)$ speculative evaluations per iteration, where $p$ is the swarm size.
Clearly it is much more desirable to have a sparse topology, where $O(np)$ is
much smaller than $O(p^2)$.  However, some functions are better optimized with
the Complete topology and the quick spread of information it entails than with
sparse topologies.  Accordingly, we use the Random topology described
in~\cite{mcnabb-cec09}, which has been shown to effectively simulate the
Complete topology.  In~\sec{results} we report the results for performing
speculative evaluation using both the Ring topology and the Random topology on
a number of common benchmark functions.

It is not trivial in some parallel architectures to determine which speculative
position was the correct next position of each particle.  First we discuss the
relatively easy case of a centralized parallel PSO algorithm with a master
computer and many slaves.  In such an architecture, the master keeps track of
all necessary information with only trivial message passing needed.  Then we
discuss the more complicated case of a distributed algorithm, where each
particle is on its own and needs to send and receive messages to and from other
particles.  Finally we discuss the further complications of a dynamic topology,
where a particle's neighbors change from one iteration to another.

@@

Upon picking the correct branch for each particle and updating the child's
personal best and neighborhood best value (from iteration $t$), the result is
the set $\nonbest{\pset}_{t+1}$, as the particles are now no longer
speculative.  What remains is to update the neighborhood best of those
particles from their neighbors (from iteration $t+1$), as above, to obtain
$\pset_{t+1}$.  That set of particles can subsequently be used to produce the
sets $\noeval{\pset}_{t+2}$ and $\noeval{\sset}_{t+3}$ (steps 1 and 2 in
\alg{centralized}), and the process repeats itself.

@@

We have already described how to use a set of messages to obtain $\p_{t+1}$.
The process is exactly the same to produce each $\n_{t+1}$, requiring the same
messages, only for the neighbor particles instead of the particle itself.
\fig{messages5} shows graphically how $\noeval{\sset}_{t+3}$ is produced.

@@

Performing speculative evaluation in PSO with a dynamic topology raises a few
issues of its own.  First, with neighbors changing each iteration, messages
that processors pass to their neighbors need to be sent to the correct
neighbors for each iteration.  In centralized algorithms and in distributed
algorithms with several rounds of message passing (the first method described
above), nothing needs to be changed.  However, in the case where processors
reconstruct information about neighboring particles from messages, those
messages need to be sent to the correct neighbors.  A particle cannot simply
send messages to its neighbors' neighbors' neighbors---it needs to send
messages to its iteration $t$ neighbors' iteration $t+1$ neighbors, and so on.
For every neighbor outward information is sent, the iteration also needs to be
incremented, as information about neighbors' neighbors is used during iteration
$t+1$, and information about neighbors' neighbors' neighbors is used to
reconstruct information about iteration $t+2$.  This method of message passing
again requires the use of random seeds if the topology is random, so that each
processor computes the same neighbors for a particle as all other processors.

An issue common to all methods of performing speculative evaluation is that of
getting available information from neighbors before generating speculative
children.  In a static topology, at iteration $t$ a particle already has all of
the information about the positions of its neighbors during iterations $1$
through $t-1$.  If the neighbor finds a better position at iteration $t$, the
particle updates its neighborhood best, but if it does not, it still has its
old neighborhood best from its neighbors for all previous iterations.

In a dynamic topology, a particle might not have information about the previous
positions of its neighbors at iteration $t$.  That means that its new
neighborhood best could come not only from its neighbors' positions at
iteration $t$, but also from their personal best from iteration $t-1$, as
neighbors' personal bests are what are used to update a particle's neighborhood
best.  That creates a problem for speculative evaluation---there are
potentially more than $2n+1$ possible next positions, increasing the amount
of work that must be done to perform the second iteration at the same time as
the first.

This is easily fixed by updating $\noeval{\p}_{t+2}$ with the currently
available information about $\noeval{\nset}_{t+2}$ before producing
$\noeval{\sset}_{t+3}$.  If a particle $\noeval{\p}_{t+2}$ updates its 
neighborhood best with the personal bests of $\noeval{\nset}_{t+2}$ before
calculating the next possible positions for $\noeval{\sset}_{t+3}$, there are
still only $2n+1$ possible next positions, and the problem is avoided.

@@

In performing speculative evaluation as we have described it, $2n+1$
speculative evaluations are done per particle, while all but one of them are
completely ignored.  It seems reasonable to try to make use of those
evaluations instead of throwing them away.  Making use of them changes the
behavior of PSO, instead of reproducing it exactly as the above method
explains, but the change could be an improvement.

@@

In our experiments we compared our speculative PSO algorithm to the standard
PSO algorithm.  At each iteration of the algorithms, we use one processor to
perform one function evaluation for one particle, be it speculative or
non-speculative.  The speculative algorithm actually performs two iterations of
PSO at each ``iteration,'' so we instead call each ``iteration'' a round of
evaluations.  For benchmark functions with very fast evaluations this may not
be the best use of parallelism in PSO.  But for many real-world applications,
the objective function takes on the order of minutes or more to evaluate; in
such cases our choice of framework is perfectly reasonable.

@@

Next we look at the function Griewank, defined by the equation $f(\Vec{x}) =
\frac{1}{4000} \sum_{i=1}^D x_i^2 - \Pi_{i=1}^D \cos\left(\frac{x_i}{\sqrt{i}}
\right) + 1$.  It is generally recommended to use the Ring topology when
optimizing the Griewank function, as Complete is prone to premature convergence
on a local optimum.  Griewank has a global optimum with a value of 0, and
sometimes the swarm finds the optimum and sometimes it doesn't.  Instead of
showing average function value at each iteration, a more accurate plot for
Griewank shows the percent of runs that have found the global optimum by each
iteration.  We show results in \fig{griewank1} for swarms of size 30 and 240
using the Ring topology.  One can see in the figure that both SEPSO algorithms
find the optimum faster than the original PSO.  However, because the swarm size
is so small, SEPSO gets stuck a little less than half of the time.
Surprisingly, SEPSO Pick Best improves the chance of success with Griewank,
while also finding the optimum faster.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank1.eps}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations, showing percent successful by each iteration instead of average
  function value.}
  \label{fig:griewank1}
\end{figure}

Because our speculative algorithm got stuck with such a small swarm size, we
ran a set of experiments on Griewank with 800 processors, giving swarms of size
100 and 800 instead of 30 and 240.  \fig{griewank2} shows the results.
\fig{griewank2} is very similar to \fig{griewank1}, except the speculative
algorithms find the global optimum essentially 100\% of the time.  Thus with a
few more processors, our algorithms find the optimum two to three times faster
than the original PSO without a significant loss of accuracy.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank2.eps}
  \caption{Function Griewank with a swarm that uses 800 processors per round of
  evaluations.}
  \label{fig:griewank2}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{rastrigin.eps}
  \caption{Function Rastrigin with a swarm that uses 240 processors per round
  of evaluations.}
  \label{fig:rastrigin}
\end{figure}

@@

We have described how parallel implementations of particle swarm optimization
can be modified to allow additional processors to increase the number of
iterations of the algorithm performed, instead of merely adding more particles
to the swarm.  We have detailed several different implementations of the
algorithm, depending on the parallel PSO framework being used.  Using our
modifications, the original PSO algorithm is exactly reproduced two iterations
at a time.  This technique requires more function evaluations per iteration
than regular PSO, but for some functions still performs better when run in a
parallel environment.  We have also described a method for making use of extra
speculative evaluations that performs incredibly well on some functions.

@@

There are some functions for which very little exploration needs to be done;
Sphere is an example of such a function.  For such functions the best use of
processors is to have a small swarm performing speculative evaluation with our
second method, where all speculative evaluations are used.

There are other functions for which there is a point where the algorithm has
done ``enough'' exploration and consistently finds the global optimum.  Adding
additional particles to the swarm is akin to increasing its capacity for
exploration.  We have seen that for a 20-dimensional Griewank function, 30
particles is not ``enough'' exploration, but 100 is.  For these functions, one
should use processors to increase the swarm size until enough exploration has
been reached, then use additional processors to increase the number of
iterations performed.

There are still other functions for which it seems there is never enough
exploration, such as the Rastrigin function.  It has been shown that up to 4000
particles there is no point at which ``enough'' exploration has been
done~\cite{mcnabb-cec09}.  For those functions the correct use of processors
cannot be easily prescribed and depends largely on the desired outcome and how
many processors are available.  It is likely that the best thing to do with
those functions is to use all processors to add particles to the swarm and do
no speculative evaluation.
